---
title: "ggvolcano"
author: "Yurii Chinenov"
date: "April 13, 2017"
output:
  html_document: default
  word_document: default
---
##ggvolcano##

*ggvolcano* produces "volcano"" plots  ($-log_{10}(p value)$ versus $-log_2(fold change)$) that is commonly used to visualize differentially expressed genes in a two group comparison. *ggvolcano* is a *ggplot*-based function. Besides a basic volcano plot with highlights "significant" diferentially-expressed genes based on user-supplied fold change and p values cutoffs, *ggvolcano* also highlights and labels additional user-supplied gene sets. `ggvolcano` relies on *ggrepel* package by Kamil Slowikowski  to perform gene labeling. (ggrepel: Repulsive Text and Label Geoms for 'ggplot2'. R package version 0.6.5. https://CRAN.R-project.org/package=ggrepel)    

##Usage##

```
ggvolcano (gene.id=NULL, p.val=NULL, fold.c=NULL, labels="none", de.l=list(),
           top.list=NULL, cutoff.p=0.01, cutoff.fc=2, logfold.c=T, FDR=T, binhex=F)
```
###Arguments###

`gene.id`    A vector of gene names

`p.val`      A vector of *p* values

`fold.c`     A vector of fold changes 

`labels`     A vector of gene names to be labeled on the plot. If `labels="auto"` top 15 genes sorted by p values will be labeled, when `labels="none"` no labels              will be shown

`de.l`       A vector or a list of vectors with gene names that will be shown on the plot as colored circles; when no names supplied genes will be colored     
             according to p value/fold change cur offs. Each element of the list `de.l` will have unique color.

`top.list`   The number of top genes in each gene set in `de.l` that will be labeled, could be combined with user supplied `label` or `label="auto"`

`cutoff.p`   *p* value cut off. By default cutoff.p=0.01 

`cutoff.fc`   fold change cutoff. By default cutoff.fc = 2
  
`logfold.c`   a logical value indicating whether to Log2 transform fold fold change values.By default `logfold.c=TRUE`

`FDR`         a logical value indicating whether *p* values need to be adjusted for multiple comparison. By default `FDR=TRUE`, otherwise `p.addjust()` will be 
              applied with `method="fdr"`

`binhex`      a logical value indicating whether to bin nearby data points to avoid overplotting. By default `binhex=FALSE`. When `binhex=T`  'ggplot2 geom_hex` 
              will be used to decrease the number of ploted objects

###Values###

#### ```gene.id``` ####

A character vector. 
```{r, echo=T, results="hide",message=FALSE}
library(readr)
toptag1_2 <- read_csv ("toptag1_2.csv")
```

```{r, echo=F}
str (toptag1_2$genes)
```
All numeric IDs (e.g. ENTREZID) will be coersed to characters.

#### ```p.val``` ####
A numeric vector of *p* values or adjusted *p* values, for example:
```{r, echo=F}
str (toptag1_2$FDR)
```

#### ```fold.c``` ####
A numeric vector of fold changes
```{r, echo=F}
str (toptag1_2$Fold.c)
```

#### Basic volcano plot ####
```{r, fig.width =9, fig.height =8 }
ggvolcano (
  gene.id=toptag1_2$genes, 
  p.val=toptag1_2$FDR,   
  fold.c=toptag1_2$Fold.c,   
  labels="none",
  de.l=list(),
  top.list=NULL,
  cutoff.p=0.01, 
  cutoff.fc=2,
  logfold.c=T,   
  FDR=T,         
  binhex=F)  
```

#### A volcano plot with automatic labeling and dot bining:###
`binhex=T,labels="auto`

To reduce the number of plotted objects *ggvolcano* optionaly maps datapoint to hexagonal bins using `geom_hex`.
All bins are intentionally colored uniformly gray rather than based on point density as I did not consider this information particularly relevant.

```{r, warning=F, fig.width = 9, fig.height = 8}
ggvolcano (
  gene.id=toptag1_2$genes, 
  p.val=toptag1_2$FDR,   
  fold.c=toptag1_2$Fold.c,   
  labels="auto",
  de.l=list(),
  top.list=NULL,
  cutoff.p=0.01, 
  cutoff.fc=2,
  logfold.c=T,   
  FDR=T,         
  binhex=T)  
```

#### A volcano plot with overplotting and custom labeling ####

```{r, warning=F, fig.width = 9, fig.height = 8}
ggvolcano (
  gene.id=toptag1_2$genes, 
  p.val=toptag1_2$FDR,   
  fold.c=toptag1_2$Fold.c,   
  labels= toptag1_2$genes[grep("^Cxcl",toptag1_2$genes)],
  de.l=list(),
  top.list=NULL,
  cutoff.p=0.01, 
  cutoff.fc=2,
  logfold.c=T,   
  FDR=T,         
  binhex=T)  
```

#### Highlighed gene subsets, overplotting control, top 5 genes in each subset and are labeled (top.list=5)  ####
```{r, warning=F, fig.width = 9, fig.height = 8}
ggvolcano (
  gene.id=toptag1_2$genes, 
  p.val=toptag1_2$FDR,   
  fold.c=toptag1_2$Fold.c,   
  labels="none",
  de.l=list(TNF_halmark,E2F_halmark,MYC_halmark),
  top.list=3,
  cutoff.p=0.01, 
  cutoff.fc=2,
  logfold.c=T,   
  FDR=T,         
  binhex=T)  
```

#### Highlighed gene subsets, overplotting control, top 5 genes in each subset and user-provided gene set are labeled (top.list=5)  ####
```{r, warning=F, fig.width = 9, fig.height = 8}
ggvolcano (
  gene.id=toptag1_2$genes, 
  p.val=toptag1_2$FDR,   
  fold.c=toptag1_2$Fold.c,   
  labels=toptag1_2$genes[grep("^Cxcl",toptag1_2$genes)],
  de.l=list(TNF_halmark,E2F_halmark,MYC_halmark),
  top.list=3,
  cutoff.p=0.01, 
  cutoff.fc=2,
  logfold.c=T,   
  FDR=T,         
  binhex=T)  
```

#### Same as above but all data points shown  ####
```{r, warning=F, fig.width = 9, fig.height = 8}

ggvolcano (
  gene.id=toptag1_2$genes, 
  p.val=toptag1_2$FDR,   
  fold.c=toptag1_2$Fold.c,   
  labels=toptag1_2$genes[grep("^Cxcl",toptag1_2$genes)],
  de.l=list(TNF_halmark,E2F_halmark,MYC_halmark),
  top.list=3,
  cutoff.p=0.01, 
  cutoff.fc=2,
  logfold.c=T,   
  FDR=T,         
  binhex=F)  
```